Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-28 paper.tex
update mathematical description

Recombination is described by a Poisson process of rate $r$ per unit of length and time. 
A recombination to the 
left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
labelled lineages.\\

Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
of a diploid population of constant size. The newly formed lineage acquires the 
union of both ordered intervals.
Although coalescence is reciprocal, here we'll define a strict total order 
(see \ref{par:liks}) on 
all lineages in $X(t)$ based on their leftmost starting points. 
For any such strict total order on $X(t)$ and at any point in the process 
the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see \ref{par:liks}). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

The ARG as described here records the impact of every possible event along each 
inheritance path in full detail. We could however simplify this graph by 
throwing away the precise time information of all recombination events. This 
corresponds to removing the recombinant internal nodes within the ARG such that some of 
the internal nodes representing coalescence events now have 2 outgoing (backwards in time) 
edges. On the ARG in fig.\ref{fig:smc-unary}, the output of a simulation under the SMC, 
this corresponds to removing the blue internal nodes. Crucially however, 
we can still identify all lineages that were hit by a recombination event. Yet we 
can only constrain the time to this event to within a window defined by the age of the 
child node $c$ associated with that lineage and the minimum age of the parents of the 
edges connecting to $c$.\\

ARG simplification therefore touches upon the question of what is knowable.  
[TO DO: Add link with inference methods]\\

Many ARG reconstruction methods return a data structure that 
is in essence a series of local tries [REFs]. This does not need to be a contradiction. 
All ARGs can be represented as a series of local trees \citep{wong-2023}.
As long as we register each observed node when building the local tree that 
records the path through the ARG, then we can guarantee a one-to-one correspondence 
with the original ARG. Crucially, this will result in nodes that 
are locally unary. These are nodes that only have a single child along some of the 
intervals within the genome.\\

Recombination nodes are one such type of unary nodes. Figure \ref{fig:smc-unary} (bottom)
reveals however that all internal nodes but node 4 are unary along some part of 
the genome. These nodes all represent coalescence events. Note that the presence of these 
unary nodes allows us to reconstruct the full extent 
of each lineage going back through time as well as uniquely identify all lineages involved 
in a recombination event. Both the ARG and the sequence of 
local trees contain the same information, as long as they are simplified to the 
same degree.\\

[TO DO: do I need to expand on this?] Simplification is thus tied in with the concept of locally unary nodes.\\

Which ARG inference methods return locally unary nodes and how can they be inferred.
[TO DO] give a good explanation (using the figure) why these methods can infer the extent 
of coalescing lineages.
As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return the full this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes in the local trees 
ensure these two representations contain the same information. 
The ARG is augmented with haplotype information, coloured according to the 
samples that inherit from it. 
Note how the presence of a unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. It therefore modifies the 
left-to-right logic of the SMC whereby the floating lineage coalesces randomly
with a the remaining portion of the local tree. 
}
\label{fig:smc-unary}
\end{figure}

@@

What's left to compute the likelihood, is to take into account all events 
associated with $x$, 
the leftmost coordinate of $(c, p, X_{cp})$. This is either only the 
coalescence event involving 
$c$ and $d$. In this case $x=x_{c0}$. Or, $x$ can a new recombination break point that 
occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$ 
and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the 
edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$.
In the latter case the new lineage starting at $x$ can be given a temporary label 
$c^{\prime}$ such that with the same logic from the previous paragraph we 
can define $F(c^{\prime}, s, t) = \int_{s}^{t} I_{c^{\prime}}(u)du$.

\begin{equation}\label{eq:depth}
B_{(c, p)}(\theta) = \begin{cases}
e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c^{\prime}, s, t_{p})} ds \lambda^{I_{c^{\prime}d}(t_p)} & x=x_{c^{\prime}0}>x_{c0} \\
\end{cases}
\end{equation}

Here, the (second) exponential term gives the probability of not observing a 
coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further 
gives the probability of observing a recombination 
at time $s$ given a Poisson process along position $x$ with rate $r$. The 
time of the recombination is integrated out.
The last term gives the point probability density of the coalescence event 
that terminates the edge. The indicator function is used to resolve 
the non-reciprocal nature of coalescence events as defined here.\\

@@

\begin{figure}[!ht] \label{fig:algo}
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from  
local tree $T_1$ (1) to the next, $T_2$ (4). Edges that do not persist in $T_2$ are 
removed (2). Edges starting at the left coordinate of $T_2$ are inserted (3). 
After deletion or insertion of an edge the count vector $I$ is updated along the 
corresponding internode intervals. Prior to updating $I$ following an insertion, 
the likelihood for that edge is computed as $I$ then represents the total number of 
lineages the lineage associated with this new edge can 
coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
\end{figure}

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \texttt{tskit} library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strenghtens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
[NOTE: these options have already been implemented in the current implementation]

Although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm only implicitly assumes there presence.
The algorithm can therefore also accomodate for 
those inferred ARGs that lack unary nodes (e.g.\ \relate). 
Note that without unary nodes, our algorithm will detect a recombination event along 
both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
edges would switch parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1.
% simulated results
We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
After removing the recombination nodes the likelihoods as defined here were computed both with and 
without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
explained is higher when the unary nodes are retained.

>>>>

Recombination is described by a Poisson process of rate $r$ per unit of length and time. 
A recombination to the 
left of $x$ with $x>x_{i0}$ and $x<y_{im}$ splits a lineage $i$ into two new, uniquely 
labelled lineages. \\ % old lineage is removed from L, two new lineages are added

Any two \emph{overlapping} lineages coalesce at rate $\lambda = 1/(2N_e)$ in the case
of a diploid population of constant size. The newly formed lineage acquires the 
union of both ordered intervals. % L is updated accordingly
Although coalescence is reciprocal, here we'll define a strict total order 
(see \ref{par:liks}) on 
all lineages in $L(t)$ based on their leftmost starting points. 
For any such strict total order on $X(t)$ and at any point in the process 
the instantaneous rate of coalescence then equals $\lambda \sum_{i \neq j} I_{ij}$.

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has its own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see \ref{par:liks}). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

The ARG as described here records the impact of every possible event along each 
inheritance path in full detail. Fig.\ref{fig:smc-unary} shows one such possible 
ARG generated represented 
as both a graph (top) as well as a series of local trees (bottom). 
Note that both structures represent the same information. 
Crucial for the equivalence between both representations is the presence of
nodes that (locally) only have a single child along one or more local 
trees. Not all of these (locally) unary nodes are 
(blue, fig.\ref{fig:smc-unary}) recombinant nodes. By recording edges along 
the full extent of each lineage, coalescing nodes can be locally unary 
along particular genome intervals. Furthermore, the unary nodes enable 
us to uniquely identify all lineages that were hit by a recombination 
event, even when we simplify the ARG by removing all 
(blue, fig.\ref{fig:smc-unary}) recombinant nodes. 
In the absence of recombination nodes, a recombination can be observed 
easily enough on the local trees as a change in parent going from 
one tree to the next. This simplification step 
does mean however that we lose precise time information on the 
recombination events. Yet we can constrain the uncertainty to 
within a window defined by the age of the child node $c$ associated 
with that lineage and the minimum age of the parents of the 
edges connecting to $c$. This in turn is easier to visualise using the 
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).\\

ARG simplification is thus linked to the presence of unary nodes in 
the local trees and simultaneously touches upon the question of what is 
knowable. Many ARG reconstruction methods return a data structure that 
is in essence a series of local tries [REFs]. 
As highlighted, this does not need to be a contradiction. 
All ARGs can be represented as a series of local trees \citep{wong-2023}.
The important question is therefore is however about the extent by which any 
ARG reconstruction methods incorporates the available information 
on how nodes persist across these local trees.\\

[TO DO] give a good explanation (using the represented haplotypes 
in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
of coalescing lineages.
Currently, the following methods 
return the full this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.\\

[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.\\

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG generated under the SMC and its representation as a 
series of local trees (bottom). To ensure the one-to-one correspondance 
between both representations a node is recorded for each local tree whenever 
a node is encoundered along its ancestry path in the graph representation. 
This results in nodes that (locally) have only a single child, and are therefore 
(locally) unary. Blue (unary) nodes represent recombinant nodes. 
The blue arrows shows the familiar left-to-right logic of the SMC whereby 
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. 
Both ARG representations can be simplified by removing the recombinant nodes. 
In that case the information we have on the time to each of these events is 
limited to the windows (blue shaded array) delimited by by the age of 
the child node $c$ associated with that lineage and the minimum age of 
the parents of the edges connecting to $c$. 
 [TO DO: wanted to use the 
added in haplotype information to explain why we can actually \textit{infer} the 
extent of each lineage, and thus create unary nodes in the first place.]
}
\label{fig:smc-unary}
\end{figure}

@@

What's left to compute the likelihood of the focal edge, is to take into account 
all events associated with $x$, 
the leftmost coordinate of $(c, p, X_{cp})$. 
This is either only the 
coalescence event involving 
$c$ and $d$. In that case $x=x_{c0}$ and $c$ did not coalesce until $t_p$. 
Alternatively, $x$ can be a new recombination break point that 
occurred on the lineage represented by $c$ at some time $s$ between $t_{c}$ 
and $\min(t_p, t_{p^{\prime}})$, where $p^{\prime}$ is the parent of the 
edge $(c, p^{\prime}, X_{cp^{\prime}})$ ending at $x$. The new lineage formed 
by the recombination event and starting at $x$ can be given a temporary label 
$c_s$. In both cases, the cumulative hazard of a lineage $a$ coalescing 
between time $s$ and $t$ is given by $F(c, s, t) = \int_{s}^{t} I_{c}(u)du$ 
such that:

\begin{equation}\label{eq:depth}
B_{(c, p)}(\theta) = \begin{cases}
e^{-\lambda F(c, t_c, t_p)} \lambda^{I_{cd}(t_p)} & x=x_{c0} \\
\int_{t_c}^{t_{p} \wedge t_{p^{\prime}}} r e^{-rs} e^{-\lambda F(c_s, s, t_{p})} ds \lambda^{I_{c_{s}d}(t_p)} & x=x_{c_{s}0}>x_{c0} \\
\end{cases}
\end{equation}

Here, the (second) exponential term gives the probability of not observing a 
coalescence event before $t_p$. In case of a recombination, $re^{-rs}$ further 
gives the probability of observing a recombination 
at time $s$ given a Poisson process along position $x$ with rate $r$. The 
time of the recombination is integrated out.
The last term gives the point probability density of the coalescence event 
that terminates the edge. The indicator function is used to resolve 
the non-reciprocal nature of coalescence events as defined in \ref{def:coal}.\\

@@

\begin{figure}[!ht] \label{fig:algo}
\centering
\includegraphics[width=\textwidth]{figures/ts_algo_2rows.png}
\caption{Description of algorithm: Moving along the genome we transition from  
local tree $T_1$ to the next, $T_2$ in three steps. 
Edges that do not persist in $T_2$ are 
removed first. Edges starting at the left coordinate of $T_2$ are inserted. 
After deletion or insertion of an edge the count vector $I$ is updated along the 
corresponding internode intervals. Following an insertion, and prior to updating $I$, 
the likelihood for that edge is computed as $I$ then represents the total number of 
lineages the lineage associated with this new edge can 
coalesce with. By moving along the genome in this way, each edge gets visited exactly once.}
\end{figure}

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \tskit library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strenghtens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
[NOTE: these options have already been implemented in the current implementation]

Although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm only implicitly assumes their presence.
The algorithm can therefore also accomodate for 
those inferred ARGs that lack unary nodes (e.g.\ \relate). 
Without unary nodes though, our algorithm will detect a recombination event along 
both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
edges would switch parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1.

[TO DO]: improve results sections with some good viz,
We simulated 1000 ARGs under the Hudson coalescent retaining all recombination information 
and computed the corresponding likelihood using \texttt{msprime} \citep{baumdicker_efficient_2021}. 
After removing the recombination nodes the likelihoods as defined here were computed both with and 
without the unary nodes. Figure \ref{sup:fig:vs-hudson} shows that the proportion of variation 
explained is higher when the unary nodes are retained.
